---
title: "Multivariate Statistical Methods - Lab 04"
author: "Maximilian Pfundstein (maxpf364), Hector Plata (hecpl268), Aashana Nijhawan(aasni448), Lakshidaa Saigiridharan (laksa656)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    df_print: paged
    number_sections: no
    toc: yes
    toc_float: no
---

```{r setup, include=FALSE}
library(ggplot2)
library(yacca)
library(CCA)
knitr::opts_chunk$set(echo = TRUE)
```

# Canonical correlation analysis by utilizing suit- able software

Look at the data described in Exercise $10.16$ of *Johnson, Wichern*. You may find it in the file `P10-16.DAT`. The data for 46 patients are summarized in a covariance matrix, which will be analyzed in R. Read through the description of the different R packages and functions so you may chose the must suitable one for the analysis. Supplement with own code where necessary.

```{r}

data = read.table("P10-16.DAT")
head(data)

```


## Association Between Groups

**Task:** Test at the 5 percent level if there is any association between the groups of variables.

```{r}
library(expm)

S = as.matrix(data)
S11 = as.matrix(data[1:3,1:3])
S12 = as.matrix(data[4:5,1:3])
S21 = as.matrix(data[1:3,4:5])
S22 = as.matrix(data[4:5,4:5])
n=46
p=3
q=2


test_statistics = n * log((det(S11)  * det(S22) )/det(S))
critical_calue = qchisq(0.95, 3*2)

cat("Is test_statistics greater than critical_calue: ",test_statistics > critical_calue)

```

Therefore, we reject $H_0$


## Number of Cononical Significant Variable

**Task:** How many pairs of canonical variates are significant?

```{r}


eigenValVec = eigen(solve(sqrtm(S11)) %*% S21 %*% solve((S22)) %*% S12 %*% solve(sqrtm(S11)))
sqrt(eigenValVec$values)



```


## Interpretation of the Significant Squared Canonical Correlations

**Task:** Interpret the “significant” squared canonical correlations.
**Tip:** Read section “Canonical Correlations as Generalizations of Other Correlation Coeffi- cients”.

```{r}

```


## Interpretation of Canonial Variates

**Task:** Interpret the canonical variates by using the coefficients and suitable correlations.

```{r}

```


## Suitability of the Canonical Variates as a Summary Measure

**Task:** Are the “significant” canonical variates good summary measures of the respective data sets?
**Tip:** Read section “Proportions of Explained Sample Variance”.

```{r}

```


## Opinion on the Success of the canonical analysis.

**Task:** Give your opinion on the success of this canonical correlation analysis.

```{r}

```

